### 2022-11-17 lmcs/conclusion.tex
another pass on related work

\subsubsection*{Deadlock freedom and progress}
Deadlock freedom and progress are well studied properties in the $\pi$-calculus.
For the `standard' typed $\pi$-calculus, an important line of work starts from Kobayashi's approach to deadlock freedom~\cite{kobayashi98}, where priorities are values from an abstract poset. Kobayashi~\cite{kobayashi06} simplifies the abstract poset to pairs of naturals, called \emph{obligations} and \emph{capabilities}. Padovani simplifies these further to a single natural, called a \emph{priority}~\cite{padovani14}, and adapts obligations/capabilities to session types~\cite{padovani13}.

% We can obtain deadlock freedom for the $\pi$-calculus with session types via the encoding of session types using linear types~\cite{kobayashi07,dardhagiachino12,dardha14beat} and linear $\pi$-calculus with priorities.

For the session-typed $\pi$-calculus, Dezani~\etal~\cite{dezani-ciancaglinimostrous06} guarantee progress by allowing only one active session at a time. Dezani~\cite{dezani-ciancagliniliguoro09progress} introduces a partial order on channels, similar to Kobayashi~\cite{kobayashi98}.
Carbone and Debois~\cite{carbonedebois10} define progress for session typed $\pi$-calculus in terms of a \emph{catalyser} which provides the missing counterpart to a process. Carbone~\etal~\cite{carbonedardha14} use catalysers to show that progress is a compositional form of lock-freedom and can be lifted to session types via the encoding of session types to linear types~\cite{kobayashi07,dardhagiachino12,dardha14beat}.
Vieira and Vasconcelos~\cite{vieiravasconcelos13} use single priorities and an abstract partial order to guarantee deadlock freedom in a binary session-typed $\pi$-calculus and building on conversation types.

@@

\subsubsection*{Ties with logic}
The correspondence between logic and types lays the foundation for functional programming~\cite{wadler15}.
Since its inception by Girard~\cite{girard87}, linear logic has been a candidate for a foundational correspondence for concurrent programs.
A~correspondence with linear $\pi$-calculus was established early on by Abramsky~\cite{abramsky94} and Bellin and Scott~\cite{bellinscott94}. Many years later, several correspondences between linear logic and the $\pi$-calculus with binary session types were proposed. Caires and Pfenning~\cite{cairespfenning10} propose a correspondence with dual intuitionistic linear logic, while Wadler~\cite{wadler14} proposes a correspondence with classical linear logic. Both guarantee deadlock freedom as a consequence of cut elimination.
Dardha and Gay~\cite{dardhagay18extended} integrate Kobayashi and Padovani's work on priorities~\cite{kobayashi06,padovani14} with CP, loosening its ties to linear logic in exchange for expressivity.
Dardha and P\'{e}rez~\cite{dardhaperez15extended} compare priorities \`a la Kobayashi with tree restrictions \`a la CP, and show that the latter is a subsystem of the former.
Balzer~\etal~\cite{balzerpfenning17} introduce sharing at the cost of deadlock freedom, which they restore using an approach similar to priorities~\cite{balzertoninho19}.
Carbone~\etal~\cite{CarboneMSY15,carbonelindley16} give a logical view of MPST with a generalised duality.
Caires and P\'{e}rez~\cite{CairesP16} give a presentation of MPST in terms of binary session types and the use of a \emph{medium process} which guarantee protocol fidelity and deadlock freedom. Their binary session types are rooted in linear logic.
Ciobanu and Horne~\cite{CiobanuH15} give the first Curry-Howard correspondence between MPST and BV~\cite{Guglielmi07}, a conservative extension of linear logic with a non-commutative operator for sequencing.
Horne~\cite{Horne20} give a system for subtyping and multiparty compatibility where compatible processes are race free and deadlock free using a Curry-Howard correspondence, similar to the approach in~\cite{CiobanuH15}.

@@

% \subsubsection*{Acknowledgements}
% The authors would like to thank Simon Fowler, April Gon\c{c}alves, and Philip Wadler for their comments on the manuscript.

>>>>

\subsubsection*{Deadlock freedom and progress}
Deadlock freedom and progress are well studied properties in the $\pi$-calculus.
For the `standard' typed $\pi$-calculus, an important line of work starts from Kobayashi's approach to deadlock freedom~\cite{kobayashi98}, where priorities are values from an abstract poset. Kobayashi~\cite{kobayashi06} simplifies the abstract poset to pairs of naturals, called \emph{obligations} and \emph{capabilities}. Padovani simplifies these further to a single natural, called a \emph{priority}~\cite{padovani14}, and adapts obligations/capabilities to session types~\cite{padovani13}. Later work by Kobayashi and co-authors \cite{GiachinoKL14,kobayashilaneve17} addresses deadlock detection for a value-passing CCS (predecessor of the $\pi$-calculus) where the number of nodes is arbitrary, namely modelling unbounded networks. The authors define an inference algorithm for their type system which guarantees deadlock freedom for these more complex kinds of communication networks.

For the session-typed $\pi$-calculus, Dezani~\etal~\cite{dezani-ciancaglinimostrous06} guarantee progress by allowing only one active session at a time. Dezani~\cite{dezani-ciancagliniliguoro09progress} introduces a partial order on channels, similar to Kobayashi~\cite{kobayashi98}.
Carbone and Debois~\cite{carbonedebois10} define progress for session typed $\pi$-calculus in terms of a \emph{catalyser} which provides the missing counterpart to a process. Carbone~\etal~\cite{carbonedardha14} use catalysers to show that progress is a compositional form of lock-freedom and can be lifted to session types via the encoding of session types to linear types~\cite{kobayashi07,dardhagiachino12,dardha14beat,dardhaetal17}.

Vieira and Vasconcelos~\cite{vieiravasconcelos13} use single priorities and an abstract partial order to guarantee deadlock freedom in a binary session-typed $\pi$-calculus and building on conversation types.
Coppo \etal \cite{coppoetal13,coppoetal16} present a deterministic, sound and complete, and compositional type inference algorithm for an \emph{interaction type system}, which is used to guarantee global progress for processes in a calculus based on asynchronous as well as dynamically interleaved and interfered multiparty sessions. The interaction type system allows to infer \emph{causalities}---much in the line of Kobayashi's priorities ---of channels, thus guaranteeing that session-typed processes do not get stuck at intermediate stages of their sessions.

@@

\subsubsection*{Ties with logic}
The correspondence between logic and types lays the foundation for functional programming~\cite{wadler15}.
Since its inception by Girard~\cite{girard87}, linear logic has been a candidate for a foundational correspondence for concurrent programs.
A~correspondence with linear $\pi$-calculus was established early on by Abramsky~\cite{abramsky94} and Bellin and Scott~\cite{bellinscott94}. Many years later, several correspondences between linear logic and the $\pi$-calculus with binary session types were proposed. Caires and Pfenning~\cite{cairespfenning10} propose a correspondence with dual intuitionistic linear logic, while Wadler~\cite{wadler12} proposes a correspondence with classical linear logic. Both works guarantee deadlock freedom as a consequence of adopting cut elimination.
Building on a previous work \cite{cairespfenning10}, Toninho \etal \cite{toninhoetal13} present a Curry-Howard correspondence between session types and linear logic for functional language via linear contextual monads, which are first-class values, thus giving rise to a higher-order session typed  language. In addition to the more standard results, the authors also prove a global progress theorem.
Dardha and Gay~\cite{dardhagay18extended} define Priority CP by integrating Kobayashi and Padovani's work on priorities~\cite{kobayashi06,padovani14} with CP, loosening its ties to linear logic in exchange for expressivity. However, they show how PCP can be also \emph{viewed as} a an extension of linear logic, which they call Priority Linear Logic (PLL), and uses mix and cycle rules as opposed to the cut rule.
Dardha and P\'{e}rez~\cite{dardhaperez15extended,dardhaperez15,DardhaP22} compare priorities \`a la Kobayashi with tree restrictions \`a la CP, and show that the latter is a subsystem of the former. In addition, they give a detailed account of comparing several type systems for deadlock freedom spanning across session types, linear logic, and linear types.
Carbone~\etal~\cite{CarboneMSY15,carbonelindley16} give a logical view of MPST with a generalised duality.
Caires and P\'{e}rez~\cite{CairesP16} give a presentation of MPST in terms of binary session types and the use of a \emph{medium process} which guarantee protocol fidelity and deadlock freedom. Their binary session types are rooted in linear logic.
Ciobanu and Horne~\cite{CiobanuH15} give the first Curry-Howard correspondence between MPST and BV~\cite{Guglielmi07}, a conservative extension of linear logic with a non-commutative operator for sequencing.
Horne~\cite{Horne20} give a system for subtyping and multiparty compatibility where compatible processes are race free and deadlock free using a Curry-Howard correspondence, similar to the approach in~\cite{CiobanuH15}.
Balzer~\etal~\cite{balzerpfenning17} introduce sharing at the cost of deadlock freedom, which they restore using \emph{worlds}, an approach similar to priorities~\cite{balzertoninho19}.
Staying on sharing, Rocha and Caires \cite{rochacaires21} introduce an imperative feature, that of shared mutable states into a functional language based on Curry-Howard correspondence with linear logic. Their type system is thus able to capture programs which were not possible in previous works. The authors prove extensive technical results, including session fidelity, progress, confluence and
normalisation.
Lastly, Jacobs \etal \cite{JacobsBK22a} present a novel technique to guarantee deadlock freedom based on the notion of connectivity graph, which is an abstract representation of the topology of concurrent systems, and separation logic used to substructurally treat connectivity graph edges and labels.

@@

\subsubsection*{Acknowledgements}
The authors would like to thank the anonymous reviewers for their detailed feedback, which helped produce a more complete and polished work.
Also, the authors would like to thank Simon Fowler, April Gon\c{c}alves, and Philip Wadler for their comments on the manuscript.

### 2022-11-17 lmcs/priority-gv.tex
another pass on related work

\emph{Actions} are terms which perform communication actions and which synchronise between two threads.
\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
Progress for the term language is standard for GV, and deviates from progress for linear $\lambda$-calculus only in that terms may reduce to values or \emph{ready terms}:
\begin{definition}[Actions]%
  \label{def:pgv-actions}
  A~term \emph{acts on} an endpoint $\tm{x}$ if it is $\tm{\send\;\pair{V}{x}}$, $\tm{\recv\;{x}}$, $\tm{\close\;{x}}$, or $\tm{\wait\;{x}}$. A~term is an \emph{action} if it acts on some endpoint $\tm{x}$.
\end{definition}
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$.
\end{definition}
\begin{lemma}[Progress, $\tred$]%
  \label{lem:pgv-open-progress-terms}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\ty{\Gamma}$ contains only session types, then:
  \begin{enumerate*}[label= (\alph*) ]
  \item $\tm{M}$ is a value;
  \item $\tm{M}\tred\tm{N}$ for some $\tm{N}$; or
  \item $\tm{M}$ is ready.
  \end{enumerate*}
\end{lemma}

>>>>

\emph{Actions} are terms which perform communication actions and which synchronise between two threads.
\begin{definition}[Actions]%
  \label{def:pgv-actions}
  A~term \emph{acts on} an endpoint $\tm{x}$ if it is $\tm{\send\;\pair{V}{x}}$, $\tm{\recv\;{x}}$, $\tm{\close\;{x}}$, or $\tm{\wait\;{x}}$. A~term is an \emph{action} if it acts on some endpoint $\tm{x}$.
\end{definition}

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to \emph{live} processes  introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and \emph{poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$.
\end{definition}

\emph{Progress} for the term language is standard for GV, and deviates from progress for linear $\lambda$-calculus only in that terms may reduce to values or \emph{ready terms}.

\begin{lemma}[Progress, $\tred$]%
  \label{lem:pgv-open-progress-terms}
  If $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ and $\ty{\Gamma}$ contains only session types, then:
  \begin{enumerate*}[label= (\alph*) ]
  \item $\tm{M}$ is a value;
  \item $\tm{M}\tred\tm{N}$ for some $\tm{N}$; or
  \item $\tm{M}$ is ready.
  \end{enumerate*}
\end{lemma}
